---
title: "Question 2 methods report"
date: "June 29, 2022"
author:
  - June Mi Jennifer Kang^[Georgia State University & Georgia Institute of Technology]
  - Meng Ye^[Georgia State University]
  - Andrew Heiss^[Georgia State University]
output: 
  pdf_document: 
    latex_engine: xelatex
    toc: yes
    toc_depth: 3
    fig_caption: yes
bibliography: bibliography.bib
link-citations: yes
header-includes:
    - \usepackage{setspace}
    - \usepackage{float}
    - \floatplacement{figure}{H}
    - \usepackage{xcolor}
    - \usepackage{parskip}
---

\onehalfspacing
\setlength{\parskip}{12pt}

# Data challenge question

Question 2:How did fare subsidies affect long term enrollment in transit programs?

- Q2~A~: Effect of different levels of incentives on longer-term loading of value and passes
- Q2~B~: Effect of different levels of incentives on re-enrollment in ORCA LIFT
- Q2~C~ and Q2~D~: Effects of subsidized passes on loading and enrollment

# Research design: Inverse Probability Weighting (IPW)

We use inverse probability weighting to adjust for confounding in the relationship between incentives and loading and enrollment.

## Directed Acyclic Graphs (DAGs)

In identifying the causal relationship between treatment and outcome, a particular type of causal graph called Directed Acyclic Graphs (DAGs) can become incredibly helpful. DAGs are instrumental in identifying confounding factors that bias the relationship between treatment and outcome variables by opening up backdoors [@pearl_book_2018]. For instance, refer to a simplified, hypothetical version of DAG for the first part of Question 2, asking the following question: what are the causal effects of different levels of incentives on longer-term ridership?
\

```{r load-libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)  # ggplot, dplyr, %>%, and friends
library(ggdag)  # Make DAGs with ggplot
library(dagitty)  # Do basic DAG math
library(broom)  # For converting model output to data frames
```


```{r dag, message=FALSE, warning=FALSE, fig.cap="A Simplified Demonstrative DAG for Q2 to Explain Confounders"}

node_details <- tribble(
  ~name, ~label, ~x, ~y,
  "incentive", "Incentive", 2, 1,
  "ridership", "Long-term Ridership", 4, 1,
  "unmeasured", "Unmeasured Confounder", 1, 2,
  "income", "Income Level", 2.5, 2,
  "health", "Health Condition", 5, 2,
  "background", "Background", 2, 3,
  "edu", "Educational Attainment", 4, 3
)

node_labels <- node_details$label
names(node_labels) <- node_details$name

Q2_dag <- dagify(ridership ~ incentive + income + health + 
                          edu + background,
                        incentive ~ unmeasured + income + health, 
                        unmeasured ~ background + edu + income,
                        income ~ background + edu,
                        health ~ background + edu,
                        edu ~ background,
                        exposure = "incentive",
                        outcome = "ridership",
                        latent = "background",
                        coords = node_details,
                        labels = node_labels)

# Turn DAG into a tidy data frame for plotting
Q2_dag_tidy <- Q2_dag %>% 
  tidy_dagitty() %>%
  node_status()   # Add column for exposure/outcome/latent

status_colors <- c(exposure = "#F16F6F", outcome = "#00BABA", latent = "grey50")

# Drawing DAG
ggplot(Q2_dag_tidy, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) +
  geom_dag_label_repel(aes(label = label, fill = status), seed = 1234,
                       color = "white", fontface = "bold") +
  scale_color_manual(values = status_colors, na.value = "grey20") +
  scale_fill_manual(values = status_colors, na.value = "grey20") +
  guides(color = FALSE, fill = FALSE) + 
  theme_dag()

```


## Adjusting for Confounders

Here, we use ggdag/daggity packages to draw a relatively simplified causal model to explain the causality behind longer-term ridership outcomes. For the first part of question 2, we are interested in the direct path between our treatment (different levels of incentives) and outcome (longer-term ridership). However, several potential backdoor confounders could have impacted both the treatment and outcome variables (degree of incentives &rarr; longer-term ridership). 

First, we observe that the outcome variable of longer-term ridership is caused by various factors, ranging from innate backgrounds/demographic characteristics to educational attainment, income level, health conditions, and unmeasured confounders. The arrow in DAG indicates causality- for instance, one's demographic characteristics have a causal impact on the individual's level of educational attainment. Consequently, both the person's income level and health condition are determined by educational attainment and demographic characteristics. Together, these factors generate an unobservable confounder that causes a backdoor in the direct link between the treatment and outcome. Hence, to isolate the causal link between incentive and longer-term ridership, it is necessary to account for the confounders of income level, health condition, and unmeasured confounders in this relatively simplified model, as shown below:

```{r adjust}
adjustmentSets(Q2_dag)
```

## Dealing with Unmeasured Confounders

Working with observational causal inference builds on a strong assumption that all unobservable factors that require adjustment are observable and measurable to be processed in the analysis. In an ideal world, we would be able to identify every single one of these backdoor paths *and* measure them to adjust for the biases in the analysis. However, in the real world, we do not have the ways to measure each of these background factors for every individual in our dataset. While some of the demographic information may be omitted for various reasons like privacy concerns, some of these factors also involve unmeasurable factors like reading people's minds to reveal their propensities, desires, likes, and dislikes. 

As represented in the node of "unmeasured confounder," the actual DAG for Q2 is far more complicated and involves various factors that need to be accounted for yet are unmeasurable. For instance, one's gender could have a causal effect on a confounder of drinking behaviors [@wilsnack_gender_2018] which could directly impact one's eligibility for different degrees of incentives [@pollack_substance_2002], yet these patterns are unobservable with our data. Likewise, one's income level could also have yielded another unmeasured confounder of various life opportunities, which could directly impact the level of incentives _as well as_ the values and passes in the longer term. These unmeasured confounders are problematic as they interfere with isolating the causal effects by opening up backdoor paths. The relationship often gets further complicated with the causal links _between_ these unmeasured confounders (for instance, drinking behaviors and opportunities in life), thus making it harder to isolate causality. 
\

### Using Inverse Probability Weighting (IPW)

Therefore, it requires us to utilize other techniques to adjust for confounders and biases in the model. One of the most cited statistical methods to adjust for these confounders and close the backdoors is inverse probability weighting (IPW). IPW is a popular statistical technique to estimate causal effects by adjusting for the confounders through matching the assignment to treatment [@urdinez_causal_2021]. 

In essence, IPW estimates propensity scores to weight each observation by how distinct they are. For instance, the individuals with a high probability of treatment yet do not receive one, _or_ individuals with a low probability of treatment still receive a higher weight. Under the context of our research question, IPW assigns high weights to the individuals who had the lowest chances of getting these ridership incentives based on their other characteristics but still received one _as well as_ to the people who were the most likely to receive these incentives yet did not receive one.

### Steps to IPW Method

IPW is carried out in a two-step process:

* In the first step, propensity scores are calculated for each observation using the confounders. In the first step, we utilize the observed/measurable confounders to calculate the probability of the person involved with the unobserved confounders, also known as the propensity scores for each individual. One of the most common ways to model these propensity scores is using logistic regression. 

* Once we generate the propensity scores for each individual based on the observed characteristics, we move to the second step of converting these predicted probabilities into weights. This process of incorporating the propensity scores into the causal effect estimation model essentially generates a pseudo-population of observation [@mcgowan_understanding_2019], where some individuals are weighted more than others in estimating the causal effect of the treatment. Mathematically, the following formula is used to estimate the treatment effect in this new weighted population:


$$
\frac{\text{Treatment}}{\text{Propensity}} - \frac{1 - \text{Treatment}}{1 - \text{Propensity}}
$$

We then use these weights to adjust for the unobservable confounders by re-running the regression model with inverse probability weights. 


### Continuous Treatment Variable

While this method of IPW is a valuable tool to adjust for the confounding variables, this particular analysis employs logistic regression and hence is the most ideal for binary treatments. In the case of our research question for Question 2, we are interested in the effects of _different levels of incentives_, which are not binary. The analysis is further complicated as our data also has time aspects, as we define longer-term ridership as six months and beyond. 

Hence, there is a need to modify the IPW model with binary treatments to incorporate the _continuous treatments_. Instead of the earlier formula used to display IPW for binary treatments, we instead use the following formula to incorporate the continuous treatment variable [@naimi_constructing_2014]:

$$
IPW_{continuous\ treatment} = \displaystyle \frac{f_{X}(X;\mu_{1},\sigma^{2}_1)}{f_{X \mid C}(X\mid C=c; \mu_{2}, \sigma^{2}_{2})}
$$
where:

* $X$ refers to the continuous treatment variable
* $C$ refers to the confounding variables
* Both functions refer stand for a probability density function with a variance of $\sigma^{2}$ mean of $\mu$, whereas...
* Function in the numerator stands for the probability distribution with the treatment variable only, and
* The denominator function stands for the probability distribution of treatment _with_ the confounders.


# Overall Research Design

Based on the background information provided by the KCM, we identified the following three research focuses. We have deconstructed the primary research question into these three so that each focus has a treatment variable and an outcome variable for analysis.

* __\colorbox{purple!50}{Q2-A}__
	+ Effect of different levels of incentives* on longer-term loading of value and passes
* __\colorbox{pink}{Q2-B}__
	 + Effect of different levels of incentives* on re-enrollment in ORCA LIFT
* __\colorbox{teal!50}{Q2-C}__
	+ Effects of subsidized passes through RCTs on these two outcome variables
Note that the different levels of incentives here correspond to values such as no value, \$10, \$15, \$20, \$30, \$50, and various passports. Also, question 2 itself primarily consists of Q2-A and Q2-B. 

Q2-C is formulated to further examine the impact of subsidized passes given through RCTs. This particular question was ruled out from the main body of question 2, given the difficulty with identifying the control group in our data with non-participants getting \$10 loaded onto their cards as well. Nonetheless, we attempt to tackle this question with the aim of analyzing the general pattern in the effect of the subsidized passes given through RCTs by treating all individuals with \$10 as the controlled group.
\

##  Variable Identification Strategies

### Independent Variable Identification

* __Treatment Variable for \colorbox{purple!50}{Q2-A} and \colorbox{pink}{Q2-B}__
\
In analyze the research questions for Q2-A and Q2-B, we need to first identify the independent variable. Independent variable for both questions is the same--it is the different levels of incentives. Hence, we need a variable collapsed to display different levels of incentives, ranging from \$0, \$10, \$15, \$20, \$30, \$50, shorter passes, and subsidized annual pass.

  + To achieve this, we first pulled the variable 'Initial Load' from the dataset _Registry_, displaying values of \$0, \$10, \$15, \$20, \$30, \$50, \$70, Misc. Pass, Monthly Pass,Passport, Subsidized Annual Pass, and missing values (NAs). We also pulled the variable 'Study Card' from the same dataset, displaying the phases for the treatment.

* __Treatment Variable for \colorbox{teal!50}{Q2-C}__
\
The challenge for Q2-C was that we were unable to distinguish the control from from the treatment group since \$10 given in regular ORCA LIFT is not differentiable from \$10 control group in RCT. As previously described, we proceed by making an assumption that all individuals who received \$10 in their initial load as the control group for Q2-C. Noting its limitations, this is an an attempt to observe the general patterns, magnitude, and direction of the coefficient to make an informed estimation about the effect of the RCT. For this control variable, we then pull any individuals with initial load of \$10 from the Initial Load variable under the _Registry_ data.
\

### Dependent Variable Identification

* __Outcome Variable for \colorbox{purple!50}{Q2-A}__
\
For the outcome variable for Q2-A, we are interested in analyzing the _longer-term loading of values and passes_. By definition, the longer-term refers to 6-month period and beyond since the date card was issued or initial load of values/passes were initiated. 

    + There are two different ways to approach whether the longer-term ridership has increased or decreased. One way is by examining _frequency_ of reloading. For this, we utilize the 'DateIssused' variable from the _Registry_ data as well as 'week' variable from _Sales_. Another approach to examining ridership pattern is through values loaded onto the cards. For this, we focus on the variables 'Txn Desc' (i.e. type of product purchased, whether purse/load or pass). 'loadings', and 'Amount' from _Sales_ data.

* __Outcome Variable for \colorbox{pink}{Q2-B}__
\
For Q2-B, we are interested in the effects of different levels of incentives on re-enrollment. In the data _Registry_, re-enrollments are identified through the variable 'card_ID' or the anonymized card ID unique to each individual. Re-enrollment is indicated through the number following the dash after the card number. Hence, we separate the number after the dash to push forward with our analysis.

* __Outcome Variable for \colorbox{teal!50}{Q2-C}__
\
For Q2-C, the treatment variable here is whether the individual was selected into the Randomized Controlled Trial (RCT) or not to receive the subsidized annual pass. Hence, from the _Registry_ dataset, we pulled the same variables of 'Initial Load' as well as 'Study Card' to extract the independent variable for analysis.


### Control Variables

* __From the Question 2 Data__

In addition to these key treatment and outcome variables, there are several control variables provided within the data, attached to each observation. The key variables are the variables 'Age," 'RaceDesc,' and 'LanguageSpoken' from the _Registry_ data. In addition, we also isolated 'CardIssuingAgency' to examine the location of these issuing agencies/offices as a potential instrumental variable to control for any potential endogeneity issue. We also extract the boarding information from _Boarding_ data to represent the number of boarding made on different transit services on a given week.

* __Connecting to ACS Data__

While ACS data was primarily highlighted for the Question 1 of this data challenge, our team thought it would be beneficial to further expand the pool of control variables by connecting Question 2 data to ACS. In the data _Registry_ is FIPS code of home address for each individual. Using this code, we are able to connect the home address of each individual and connect to ACS 5-year aggregate data, which would open up doors for a whole host of control variables of interest. We use 2019 ACS block group-level data to adjust for demographic and socioeconomic characteristics of riders (see `R/get_census_data.R` for details).


## Data Processing Flow 

* __Data Linking/Cleaning Strategies__

First, we merged all these identified variables into a single file using the common variable of "card_id", connecting across the three datasets of _Registry_, _Sales_, and _Boardings_.

One of the key challenges with cleaning the data was how to format the different levels of incentives. On the surface, they may seem as if in ordinal manner; however, the increments between values vary from one another, making it harder to compare. For instance, a jump from \$10 to \$15 dollar isn't quite the same as the jump from \$30 to the next closet value of \$50. Another challenge was conceptualizing how to order different types of passes. Hence, we have reduced these categories into the following groups to make a meaningful comparison across them: _Nothing, \$0, \$10, \>$10, shorter passes, and subsidized annual passes_


## Results and Discussion

In addition to several descriptive analyses, we performed inverse probability weighting (IPW) to analyze the effects of each treatment on the outcome variable.

* __Outcome Variable for \colorbox{purple!50}{Q2-A}__
\
```{r Q2-A-1, fig.align = 'center', out.width = "70%", echo=FALSE}
knitr::include_graphics("output/model_q2a_1_comparisons.pdf")
```
   + We find a puzzling result--when compared to no incentives, each type of incentive causes a decrease in the probability/proportion of reloading 6 months after and beyond.
   
  + Similarly, our results show that different levels of incentives through the program yield unexpected causal impacts on amount loaded:
   
   
```{r Q2-A-2, fig.align = 'center', out.width = "70%", echo=FALSE}
knitr::include_graphics("output/model_q2a_2_comparisons.pdf")
```

   + We see a similar pattern on the effects of different incentive levels on the card loadings. When calculated at the average count of card loadings, we see a reducing pattern compared to no incentive at all:
   
```{r Q2-A-3, fig.align = 'center', out.width = "70%", echo=FALSE}
knitr::include_graphics("output/model_q2a_3_comparisons.pdf")
```

* __Outcome Variable for \colorbox{pink}{Q2-B} and \colorbox{teal!50}{Q2-C}__ 
\
As demonstrated in a separate paper for coding, we did not yield meaningful results for Q2-B and Q2-C, despite numerous attempts to tackle the causal analysis.

\newpage

# Limitations 
While we did extend our data analysis period to the entirety of the data to include as many observations as possible, the data inherently included the ridership data during the pandemic. Even though the pandemic has made a large impact across all cities in America, Seattle and hence therein King County was the place that experienced significant changes--including restrictions on business and having face-to-face classes. Such restrictions inherently reduced ridership of the primary population to King County Metro who used the transportation system to commute to work/schools on daily basis [@emc_research_king_2020]. Such impact from the pandemic naturally caused noises in our data that would have interfered with the analysis results.

# Policy Implications
Therefore, we encourage the KCM to further push with its evidence-based policy making initiatives. The pandemic is still evidently ongoing and impacting everyday lives, hence the riders' behaviors. The KCM has already implemented two RCTs in effort to investigate the causal impacts of the subsidies on ridership, in addition to this CISIL data challenge. Continued support for such commitment would allow KCM to gather data with increasingly reduced impacts from the pandemic, eventually allowing sufficient data to observe the impacts of its various incentive programs.
\

# References
  
  
